A Bioconductor Workflow for Processing and Analysing Spatial Proteomics Data

### 2016-06-02 bioc-workflow.Rmd
data processing section

```{r env, echo=FALSE}
library("BiocStyle")
suppressPackageStartupMessages(library("MSnbase"))
suppressPackageStartupMessages(library("pRoloc"))
```

@@

# Reading and handling mass-spectrometry based proteomics data

@@

@@

* normalisation
* missing data
* combining acquisistions

Before combination, the two replicates were separately normalised by
sum across the 10 channels (i.e. such that the sum of each protein's
intensity is 1), for each replicate respectively. Normalisation is an
essential part of data processing and several methods are available in
`r Biocpkg("MSnbase")`. The normalisation desired in this specific
case would be obtained with the a call to the `normalise` method.

The combination of data can also be performed effectively in 
`r Biocpkg("MSnbase")` as detailed in the dedicated *Combining MSnSet
instances* section of the `r Biocpkg("MSnbase")`
[tutorial vignette](http://bioconductor.org/packages/release/bioc/vignettes/MSnbase/inst/doc/MSnbase-demo.pdf).

>>>>

```{r env, echo=FALSE}
library("BiocStyle")
suppressPackageStartupMessages(library("MSnbase"))
suppressPackageStartupMessages(library("pRoloc"))
suppressPackageStartupMessages(library("pRolocdata"))
```

@@

# Reading and processing spatial proteomics data

@@

## Importing data

@@

### Normalisation

There are two aspects related to data normalisation that are relevant
to spatial proteomics data processing. The first one focuses on
reducing purely technical variation between fractions without
affecting biological variability (i.e. the shape of the quantitative
profiles) too much, as generally focused on. This normalisation will
depend on the underlying quantitative technology and the experimental
design, and will not be addressed in this workflow. In addition, and
more specific to spatial proteomics data, it is fundamental to scale
all the organelle-specific profiles into a same intensity interval
(typically 0 and 1) by, for example, diving each intensity by the sum
of the intensities for that quantitative feature, as illustrated
below.

```{r normsum, eval = FALSE}
lopit2016 <- normalise(lopit2016, method = "sum")
```

This transformation of the data assures to cancel the effect of the
absolute intensities of the quantitive features along the rows, to
focus subsequent analyses on the relative profiles along the
sub-cellular fractions.

The same `normalise` function (or `normalize`, both spellings are
supported) can also be applied in the first case described
above. Different normalisation methods such as mean or median scaling,
variance stabilitation or quantile normalisation, to cite a few, can
be applied to accomodation different needs.

Before combination, the two replicates in the `lopit2016` data that we
read into R were separately normalised by sum across the 10 channels
for each replicate respectively. We can verify this by summing each
rows for each replicate:

```{r normcheck}
summary(rowSums(exprs(lopit2016[, lopit2016$replicate == 1])))
summary(rowSums(exprs(lopit2016[, lopit2016$replicate == 2])))
```
We see that some features do not add up exactly to 1 due to rounding
errors after exporting to intermediate files. These small deviations
do not bear any consequences here.

### Combining acquisitions

The spreadsheet that was used to create the `lopit2016` MSnSet
included two replicates. We also provide individual replicates in the
`r Biocexptpkg("pRolocdata")` package. Below, we show how to combine
`MSnSet` objects and, subsequently, how to filter and hangle missing
values. We start by loading the experimental data package and the
equivalent replicates using the `load` function.

```{r reps4combine}
library("pRolocdata")
data(hyperLOPIT2015ms3r1)
data(hyperLOPIT2015ms3r2)
```

To list all data distributed with a package, one can use pass the
package name as input to the `data` function. For instance

```{r pkgdata, eval=FALSE}
data(package = "pRolocdata")
```

will list the `r nrow(data(package = "pRolocdata")$results)` datasets
that are available in `r Biocexptpkg("pRolocdata")`.

Combining data is performed with the `combine` function. This function
will inspect the feature and sample names to identify how to combine
the data. As we want our replicates to be combined along the columns
(same proteins, different sets of fractions), we need to assure that
the respective sample names differ.

```{r samplenames}
identical(sampleNames(hyperLOPIT2015ms3r1), sampleNames(hyperLOPIT2015ms3r2))
hyperLOPIT2015ms3r1 <- updateSampleNames(hyperLOPIT2015ms3r1, 1)
hyperLOPIT2015ms3r2 <- updateSampleNames(hyperLOPIT2015ms3r2, 2)
sampleNames(hyperLOPIT2015ms3r1)
sampleNames(hyperLOPIT2015ms3r2)
```

In addition, to matching names, the content of the feature metadata
for identical feature annotations must match exactly across the data
to be combined. In particular for these data, we expect the same
proteins in each replicates to be annotated with the same UniProt
entry names and descriptions, but not with the same coverage of number
of peptides or peptide-spectrum matches (PSMs).

```{r fvarnames}
fvarLabels(hyperLOPIT2015ms3r1)
fvarLabels(hyperLOPIT2015ms3r2)
```

Below, we update the replicate specific feature variable names and
remove the shared annotation.

```{r r fvarnames2}
fvarLabels(hyperLOPIT2015ms3r1)[3:5] <- paste0(fvarLabels(hyperLOPIT2015ms3r1)[3:5], 1)
fvarLabels(hyperLOPIT2015ms3r2)[3:5] <- paste0(fvarLabels(hyperLOPIT2015ms3r2)[3:5], 2)
fData(hyperLOPIT2015ms3r2) <- fData(hyperLOPIT2015ms3r2)[3:5]
```

We can now combine the two experiments into a single `MSnSet`:

```{r combine}
hyperLOPIT <- combine(hyperLOPIT2015ms3r1, hyperLOPIT2015ms3r2)
hyperLOPIT
```

More details above combining data are given in the dedicated
*Combining MSnSet instances* section of the `r Biocpkg("MSnbase")`
[tutorial vignette](http://bioconductor.org/packages/release/bioc/vignettes/MSnbase/inst/doc/MSnbase-demo.pdf).

### Missing data

Missing data are a recurrent issue in mass spectrometry applications,
and should be addressed independently of this workflow
[@Lazar:2016]. In [@Gatto:2014b], we have described how a high content
in missing values in spatial proteomics data and their inappropriate
handling leads to a reduction of sub-cellular resolution.

```{r namap}
image2(is.na(hyperLOPIT), col = c("black", "white"))
```

## Replication

See issue #6.
